Testing the correlates of vast vertebrae number disparity in Australian blindsnakes

### 2024-10-29 manuscript.Rmd
add phylogenetic signal

We examined 341 specimens representing 46 typhlopid lineages including 42 of 47 formally described *Anilios*, two genetically distinct undescribed *Anilios* [@tiatragulAustralia2023;uetzQuarter2021], and their close relatives *Ramphotyphlops multilineatus* (Schlegel, 1839) and *Acutotyphlops subocularis* (Waite, 1897). Our sample size per species ranged from 1 to 25 specimens (median = 7 specimens). We sourced specimens from the South Australian Museum (SAMA), Western Australian Museum (WAM), Queensland Museum (QM), Museum and Art Gallery of the Northern Territory (MAGNT), and a private collection. Where possible, we chose relatively large non-fragile specimens from both males and females with well-preserved heads for X-ray radiography. Sex for each specimen was determined by inspecting the gonads or assigned based on relative tail lengths following @tiatragulMorphological2024.

@@

We collected body size measurements and vertebral counts from museum specimens (**Table S1**). Body measurements including snout-to-vent length (SVL), tail lengths (TL) and midbody diameter (MBD) were obtained from @tiatragulMorphological2024. SVL and TL were measured to the nearest 1 mm on a ruler. MBD was measured using a digital vernier calliper (ABS Coolant Proof Caliper, Mitutoyo, Japan) to the nearest 0.1 mm. Total body length (TBL) is the sum of SVL and TL. To quantify body shape for each specimen, we calculated the aspect ratio — also referred to as elongation ratio [@wardConvergent2004] — which is the quotient of TBL and MBD.

@@

All statistical analyses were performed in R Statistical Environment v 4.4.0 [@rcoreteamLanguage2024]. To test for evolutionary correlations between the number of vertebrae and body size and proportions, we used phylogenetic generalised least squares (PGLS) regression under a model of Brownian Motion implemented in the R package *`geomorph`* v.4.0.6 [@adamsGeomorph2023]. First, we tested for pleomerism [@lindseyPleomerism1975] across 37 species of *Anilios* blindsnakes. We did this by regressing the species maximum number of total vertebrae against maximum TBL. Next, we tested whether the species mean aspect ratio is correlated with mean vertebra ratio (mean number of vertebrae/mean TBL) and mean number of vertebrae. Finally, we tested if variation in vertebra ratio and number of vertebrae is correlated with maximum soil bulk density and mean annual temperature. In all PGLS analyses, statistical significance was assessed by permutation, using 1000 iterations. To visualise the relationship between number of vertebrae and intracolumnar vertebral length for each of the representative species, we plotted the mean vertebra size (from three repeated measurements) against the vertebra number. Each vertebra profile was overlaid with a fitted 4° polynomial curve [e.g., @sherrattMacroevolution2022].

@@

Total number of vertebrae varies greatly among the Australo-Papuan blindsnakes in this study (Fig. \@ref(fig:fig-tree-boxplots)), ranging from 146 in *A. tovelli* to 400 in *A. grypus* (**Table S1**). Aspect ratios also vary greatly, ranging from 20.8 in the robust-and-wide *A. pinguis* to 169.8 in the narrow-and-long *A. longissimus* (**Table S2**). Intraspecific variation is also present in both aspect ratio and total number of vertebrae (Fig. \@ref(fig:fig-tree-boxplots)).

@@

For the Australian blindsnake genus *Anilios*, we found a weakly positive but non-statistically significant relationship between maximum number of total vertebrae and maximum TBL (PGLS, slope = 0.28, intercept = 256, R^2^ = 0.02, F = 0.74, Z = 0.30, P = 0.41), indicating an absence of pleomerism in Australian blindsnakes (Fig. \@ref(fig:fig-tbl-vert) and **Fig S1**). This suggests that maximum length is achieved by elongating each vertebra rather than by adding additional vertebrae. Many species have more vertebrae than expected for their size. For example, *A. grypus* "West" is an intermediate length (relative to other species) but has the greatest number of vertebrae, while *A. torresianus*, which has the greatest maximum TBL, has relatively few vertebrae.

@@

In a related study, @tiatragulMorphological2024 found that Australian blindsnake species with smaller absolute body size generally occupy warmer areas than larger species tend to occupy cooler regions. Here, we observe that vertebra ratio (raw number of vertebrae/SVL) is positively correlated with temperature, indicating species that have more vertebrae than expected from their body size are found in warmer places. In his work on rattlesnakes, @klauberRattlesnakes1982  noted that smaller-sized subspecies often retain the same number of vertebrae as their larger closest relative, which results in the smaller subspecies having a relatively greater vertebra ratio. An explanation for this pattern is that size is more likely to be under greater selection than vertebral number, and there is negligible disadvantage for a snake in having too many vertebrae for their body size [@lindellEvolution1994]. This makes sense when we also consider the negative relationship between vertebra ratio and maximum soil bulk density. A greater number of (shorter) vertebrae for body size may afford greater flexibility [@jayneKinematics1986;@bergmannVertebral2012;@johnsonAdaptive1955] that enables these species to navigate in pre-existing burrows or termite galleries, typical across northern Australia, without exerting force to dig their own burrow. @tiatragulMorphological2024 also found species that occupy areas with more difficult-to-penetrate soil are usually narrow-and-long rather than robust-and-wide species that occupy areas with easier-to-penetrate soil. Narrow-and-long phenotypes are common among leptotyphlopids, which are thought to use existing burrows rather than digging their own [@herrelBurrowing2021]. In this regard, the Australian typhlopid genus *Anilios* is quite exceptional among blindsnakes because there is great variation in body shape [@tiatragulMorphological2024]. Species that evolved more narrow-and-long bodies may be converging on the behaviour with other scolecophidians that have similar phenotypes (i.e., leptotyphlopids). However, more detailed species-level microhabitat observations are required to determine if these species are actually taking advantage of these environments in the ways we speculate.

>>>>

We examined 349 specimens representing 48 typhlopid lineages including 44 *Anilios* [@tiatragulAustralia2023;uetzQuarter2021], *Ramphotyphlops multilineatus* (Schlegel, 1839), *R. acuticauda* (Peters, 1878), *R. depressus* (Peters, 1880), and *Acutotyphlops subocularis* (Waite, 1897) (**Table S1**). Our sample size per species ranged from 1 to 25 specimens (median = 6; mean = 7). We sourced specimens from the South Australian Museum (SAMA), Western Australian Museum (WAM), Queensland Museum (QM), Museum and Art Gallery of the Northern Territory (MAGNT), and a private collection. Where possible, we chose relatively large non-fragile specimens from both males and females with well-preserved heads for X-ray radiography. Sex for each specimen was determined by inspecting the gonads or assigned based on relative tail lengths following @tiatragulMorphological2024.

@@

We collected body size measurements and vertebral counts from museum specimens. Body measurements including snout-to-vent length (SVL), tail lengths (TL) and midbody diameter (MBD) were obtained from @tiatragulMorphological2024. SVL and TL were measured to the nearest 1 mm on a ruler. MBD was measured using a digital vernier calliper (ABS Coolant Proof Caliper, Mitutoyo, Japan) to the nearest 0.1 mm. Total body length (TBL) is the sum of SVL and TL. To quantify body shape for each specimen, we calculated the aspect ratio — also referred to as elongation ratio [@wardConvergent2004] — which is the quotient of TBL and MBD.

@@

All statistical analyses were performed in R Statistical Environment v 4.4.0 [@rcoreteamLanguage2024]. To test for evolutionary correlations between the number of vertebrae and body size and proportions, we used phylogenetic generalised least squares (PGLS) regression under a model of Brownian Motion implemented in the R package *`geomorph`* v.4.0.6 [@adamsGeomorph2023]. First, we tested for pleomerism [@lindseyPleomerism1975] across 37 species of *Anilios* blindsnakes. We did this by regressing the species maximum number of total vertebrae against maximum TBL. 
We also calculated the degree of phylogenetic signal from these traits separately using the "physignal.z" function in `geomorph`.
Next, we tested whether the species mean aspect ratio is correlated with mean vertebra ratio (mean number of vertebrae/mean TBL) and mean number of vertebrae. Finally, we tested if variation in vertebra ratio and number of vertebrae is correlated with maximum soil bulk density and mean annual temperature. In the phylogenetic signal and PGLS analyses, statistical significance was assessed by permutation, using 1000 iterations. To visualise the relationship between number of vertebrae and intracolumnar vertebral length for each of the representative species, we plotted the mean vertebra size (from three repeated measurements) against the vertebra number. Each vertebra profile was overlaid with a fitted 4° polynomial curve [e.g., @sherrattMacroevolution2022].

@@

Total number of vertebrae varies greatly among the Australo-Papuan blindsnakes in this study (Fig. \@ref(fig:fig-tree-boxplots)), ranging from 146 in *A. tovelli* to 400 in *A. grypus* (**Table S2**). Aspect ratios also vary greatly, ranging from 20.8 in the robust-and-wide *A. pinguis* to 169.8 in the narrow-and-long *A. longissimus* (**Table S3**). Intraspecific variation is also present in both aspect ratio and total number of vertebrae (Fig. \@ref(fig:fig-tree-boxplots)).

@@

For the Australian blindsnake genus *Anilios*, we found a weakly positive but non-statistically significant relationship between maximum number of total vertebrae and maximum TBL (PGLS, slope = 0.28, intercept = 256, R^2^ = 0.02, F = 0.74, Z = 0.30, P = 0.41), indicating an absence of pleomerism in Australian blindsnakes (Fig. \@ref(fig:fig-tbl-vert) and **Fig S1**). This suggests that maximum length is achieved by elongating each vertebra rather than by adding additional vertebrae. Many species have more vertebrae than expected for their size. For example, *A. grypus* "West" is an intermediate length (relative to other species) but has the greatest number of vertebrae, while *A. torresianus*, which has the greatest maximum TBL, has relatively few vertebrae. We found phylogenetic signal in both maximum number of total vertebrae (Z = 3.70, P = 0.001, K = 1.2, lambda = 1) and maximum TBL (Z = 3.02, P = 0.002, K = 1.0, lambda = 1).

@@

In a related study, @tiatragulMorphological2024 found that Australian blindsnake species with smaller absolute body size generally occupy warmer areas than larger species tend to occupy cooler regions. Here, we observe that vertebra ratio (raw number of vertebrae/SVL) is positively correlated with temperature, indicating species that have more vertebrae than expected from their body size are found in warmer places. In his work on rattlesnakes, @klauberRattlesnakes1982 noted that smaller-sized subspecies often retain the same number of vertebrae as their larger closest relative, which results in the smaller subspecies having a relatively greater vertebra ratio. We also observed maximum number of vertebrae in our dataset contain phylogenetic signal. An explanation for this pattern is that size is more likely to be under greater selection than vertebral number, and there is negligible disadvantage for a snake in having too many vertebrae for their body size [@lindellEvolution1994]. This makes sense when we also consider the negative relationship between vertebra ratio and maximum soil bulk density. A greater number of (shorter) vertebrae for body size may afford greater flexibility [@jayneKinematics1986;@bergmannVertebral2012;@johnsonAdaptive1955] that enables these species to navigate in pre-existing burrows or termite galleries, typical across northern Australia, without exerting force to dig their own burrow. @tiatragulMorphological2024 also found species that occupy areas with more difficult-to-penetrate soil are usually narrow-and-long rather than robust-and-wide species that occupy areas with easier-to-penetrate soil. Narrow-and-long phenotypes are common among leptotyphlopids, which are thought to use existing burrows rather than digging their own [@herrelBurrowing2021]. In this regard, the Australian typhlopid genus *Anilios* is quite exceptional among blindsnakes because there is great variation in body shape [@tiatragulMorphological2024]. Species that evolved more narrow-and-long bodies may be converging on the behaviour with other scolecophidians that have similar phenotypes (i.e., leptotyphlopids). However, more detailed species-level microhabitat observations are required to determine if these species are actually taking advantage of these environments in the ways we speculate.
